---
title: "Gov 50 Final Project"
author: "Charlotte Hodgson"
description: "My final project and milestone 1 (ideas for the final project)"
output:
  distill::distill_article:
    self_contained: false
---



## Milsetone 1: Project thoughts

I am interested in exploring data related to a variety of topics, I've listed some potential ideas below:
- in general home pricing / values in connection to another research factor - I list some ideas below: 
    -home prices in connection to public school rankings
    -housing prices closer to water source (river, ocean, lake, etc.)
    -housing prices closer to places of higher education 
    -housing prices closer to historical/protected areas 
    

## Milestone 2: Proposal Paragraph

Does a state with a higher percentage of the population having a bachelor’s degree or greater produce a higher average statewide SAT score and vice versa? In this study I plan to investigate if a state with a greater percentage of the population having a bachelor's degree or greater, thus corresponding to the percentage of 4-year college graduates, in turn causes the average SAT scores of the state to be higher in comparison to other states with lower percentages. Also, corresponding, if the higher the SAT scores than the greater bachelors degree percentage. The idea is to investigate that if more adults and parents are college graduates then the SAT-aged students in the state, potentially their children, have higher scores. This relates to the manner in which the education system is structured and also generational education trends persisting from parents to children. I hypothesize that the greater percentage of a state population that has a bachelor's degree or higher will produce a higher state SAT scoring average. Further, in turn, if the SAT scores are higher this will produce a greater percentage of the population with a bachelor's degree. The reasoning behind such a hypothesis is grounded in the generational education idea that if a parent graduated university then their child is more likely to follow those actionable steps as well, causing more emphasis to be placed on higher SAT scores. Additionally, the idea of societal norms is factorable as well since if more adults in a state are college graduates then there is potentially a greater foundation in place for college preparatory services, like SAT tutoring and practice, which can improve individual scores. Also, the wealth and education connection is another measurable factor to note as there is a notion between college graduates and higher salary jobs, which would allow them to pay for SAT preparatory practice for their children. My sample is composed of the educational population information and SAT scores of each state from the year 2022. My unit of analysis is a state (e.g., Pennsylvania). My explanatory variable of interest is the percentage of a state population with a bachelor's degree or higher. This variable will be coded in the order of least to greatest percentage of the population, a ranking. This variable will use the data from the United States Census Bureau data sets. My outcome variable is the average SAT score of each state. This will just be coded as the average SAT score, and possibly another column designated as =1 if above the average national SAT score and =0 if below the national average. This variable will be evaluated using the SAT data set of the World Population Review. If I observe trends in higher bachelor degree percentages in a state population and higher SAT scores, then my hypothesis would be supported by my review. Also, if a trend existed with states with higher bachelor degree percentages of the population and an average SAT score above the national average than that would also support my hypothesis. However, if I observe that a higher bachelor degree percentage in a state does not have a trend of higher SAT scores or SAT scores above the national average, this would provide evidence against my hypothesis. When graphing the data, if a positive trend line, or more comprehensive positive regression, is associated with the average state SAT score as the bachelor degree percentage of the population increases, that would support my hypothesis.

## Milestone 3: Data Visualization

Two scatter plots are shown below showing the distribution of one variable. The first scatter plot is of the Average SAT Scores by State, using data updated May 2023. The Scatter Plot ranks the states in the order of the lowest SAT Score to the highest SAT Score by State. The second scatter plot shows the percentage of the population that is 25 and over in each state that holds a Bachelor's Degree. The states are also ranked in order from lowest percentage to greatest percentage of bachelor's degree holders of the population 25 and over. The first plot uses data from the "World Population Review" and the second plot uses data from the "United States Census Bureau". 

```{r}
setwd("/Users/charlottehodgson/Gov 50/gov50-final-project-Hodgson")
library(readr)
library(tidyverse)

sat_data <- read_csv("SAT_SCORES.csv")
degree_percent <- read_csv("DEGREE_PERCENTAGE.csv")
drop_na(degree_percent)
```


```{r}
library(ggplot2)

ggplot(sat_data, aes(x = reorder(state, satScoresAvgTotal), y = satScoresAvgTotal)) +
  geom_point(shape = 19, size = 2) + 
  labs(title = "Average SAT Scores by State", x = "State", y = "Average SAT Scores") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.margin = margin(3, 3, 3, 1))
```
```{r}
load
```


```{r}
library(ggplot2)

ggplot(degree_percent, aes(x = reorder(State, DegreePercent), y = DegreePercent)) +
  geom_point(shape = 19, size = 2) + 
  labs(title = "Population 25 and Over with a Bachelor's Degree Percentage by State", x = "State", y = "Bachelor's Degree Percentage") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.margin = margin(1, 1, 1, 1))
```

## Milestone 4

The bar plot created ("Population 25 and Over with a Bachelor's Degree Percentage by State (Ordered by SAT Score)") merges the data sets used in the previous milestones (sat_data and degree_percent) to show the population 25 and over in each state with a bachelor's degree by percentage, but the states are ordered in lowest to highest average SAT score. This is attempting to answer my question of if a state with a higher percentage of the population having a bachelor’s degree or greater produces a higher average statewide SAT score and vice versa? The bachelor's degree percentage is shown along the y-axis and not the SAT scores, since this presented a clearer plot representation of how the SAT scores and degree percentages interact. From my hypothesis, my results would have expected to show an increase in the degree percentages as the SAT scores increase by state, however, this is not what my results show. The bar plot shows that there is not a mutual steady increase, but rather the individual categories for each state appear to have no correlation or steady increase or decline, but rather suggest no correlation between the degree percentages and SAT scores directly. However, I added a line of regression to estimate the line of best fit and see if a slope was visible. The line actually shows a slight decrease of the degree percentage as the average SAT score increases, refutung my hypothesis. 

```{r, fig.width = 8}
merged_data <- merge(sat_data, degree_percent, by.x = "state", by.y = "State", all = TRUE)

merged_data <- merged_data[order(merged_data$satScoresAvgTotal), ]

merged_data$state <- factor(merged_data$state, levels = merged_data$state)

par(mar = c(10, 5, 4, 2) + 0.1)

barplot(merged_data$DegreePercent, names.arg = merged_data$state,
        col = "hotpink", main = "Population 25 and Over with a Bachelor's Degree Percentage by State\n(Ordered by SAT Score)",
        xlab = "", ylab = "Bachelor's Degree Percent", las = 2, cex.names = 0.7) 

mtext("State (Ranked Lowest to Highest Avergae SAT Scores)", side = 1, line = 6)

fit <- lm(DegreePercent ~ as.numeric(state), data = merged_data)

abline(fit, col = "blue", lwd = 2)

```

##Final Project

Introduction: 

  Does a state with a higher percentage of the population having a bachelor’s degree or greater produce a higher average statewide SAT score and vice versa? In this study, I plan to investigate if a state with a greater percentage of the population having a bachelor’s degree or greater, thus corresponding to the percentage of 4-year college graduates, in turn, causes the average SAT scores of the state to be higher in comparison to other states with lower percentages. Also, corresponding, the higher the SAT scores the greater bachelor's degree percentage. The idea is to investigate that if more adults and parents are college graduates then the SAT-aged students in the state, potentially their children, have higher scores. This relates to how the education system is structured and also generational education trends persisting from parents to children. I hypothesize that the greater percentage of a state population that has a bachelor’s degree or higher will produce a higher state SAT scoring average. Further, in turn, if the SAT scores are higher this will produce a greater percentage of the population with a bachelor’s degree. The reasoning behind such a hypothesis is grounded in the generational trend that if a parent graduated from university then their child is more likely to follow that, causing more emphasis to be placed on higher SAT scores. Additionally, the idea of societal norms is factorable as well since if more adults in a state are college graduates then there is potentially a greater foundation in place for college preparatory services, like SAT tutoring and practice, which can improve individual scores.


Data section:

  There are two data sources for this project. One is the “Education Attainment” data set from the United States Census Bureau, the variable of population per state over the age of 25 with a bachelor's degree, and the “SAT Scores by State” data set from the World Population Review, the variable of SAT scores by state from 2023. The online United States Census Bureau data set uses information directly from the United Census Bureau in a single-platform setting. The data is found through direct response measures through survey and census models, as well as from data already collected by federal, state, or local governments. The World Population Review uses data directly shared by the College Board, which owns, develops, and publishes the SAT. 
  The independent variable, or explanatory variable, is the percentage of a state population with a bachelor’s degree or higher. This variable will be coded in the order of least to greatest percentage of the population, a ranking, measured from the surveys of statistical methodology by the United States Census Bureau. My dependent variable, or outcome variable, is the average SAT score of each state. This is measured from the data collection of test-takers and statistical modeling is used to compile the average SAT score per state from the data. Overall, statistical analysis is then used in my coding to model the comparison between these two variables. The research design is cross-sectional as the collected data is from one time versus observations over a studied period with treatments. The plot that summarizes the dependent variable, SAT scores ranked lowest to highest by state (2023), is shown below, also followed by the plot that shows the independent variable, lowest to highest ranked states by bachelor’s degree percentages (25 and up).
  
```{r}
library(ggplot2)
##create the scatter plot of the "Average SAT Scores by State"
ggplot(sat_data, aes(x = reorder(state, satScoresAvgTotal), y = satScoresAvgTotal)) +
##add the points 
  geom_point(shape = 19, size = 2) + 
#set the labels and titles 
  labs(title = "Average SAT Scores by State", x = "State", y = "Average SAT Scores") + 
##set the theme 
  theme_minimal() +
##set the angle of the text on the x-axis
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
##set the margins of the scatter plot
  theme(plot.margin = margin(3, 3, 3, 1))
```

```{r}
library(ggplot2)
##create the scatter plot of the "Population 25 and Over with a Bachelor's Degree Percentage by State"
ggplot(degree_percent, aes(x = reorder(State, DegreePercent), y = DegreePercent)) +
##add the points
  geom_point(shape = 19, size = 2) + 
#set the labels and titles
  labs(title = "Population 25 and Over with a Bachelor's Degree Percentage by State", x = "State", y = "Bachelor's Degree Percentage") + 
##set the theme
  theme_minimal() +
##set the angle of the text on the x-axis
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
##set the margins of the scatter plot
  theme(plot.margin = margin(1, 1, 1, 1))
```

Results section: 
  
Plot of main analysis
```{r}
##first merging the data sets
merged_data <- merge(sat_data, degree_percent, by.x = "state", by.y = "State", all = TRUE)

merged_data <- merged_data[order(merged_data$satScoresAvgTotal), ]

merged_data$state <- factor(merged_data$state, levels = merged_data$state)

##setting the parameters of the plot through outlining the margins to help with the spacing 
par(mar = c(10, 5, 4, 2) + 0.1)

##now creating the bar plot, with the values, color, labels, title, and the style and size of the labels 
barplot(merged_data$DegreePercent, names.arg = merged_data$state,
        col = "hotpink", main = "Population 25 and Over with a Bachelor's Degree Percentage by State\n(Ordered by SAT Score)",
        xlab = "", ylab = "Bachelor's Degree Percent", las = 2, cex.names = 0.7) 

##adding text to the bottom plot margins to reflect the title for the x-axis
mtext("State (Ranked Lowest to Highest Avergae SAT Scores)", side = 1, line = 6) 

##lastly, fitting a linear regression model to the plot and adding the regression line
fit <- lm(DegreePercent ~ as.numeric(state), data = merged_data)
abline(fit, col = "blue", lwd = 2)

```

Regression output
```{r}

##The following code presents a nicely formatted table for the regression ("fit"):

modelsummary::modelsummary(fit,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
  The bar plot created (“Population 25 and Over with a Bachelor’s Degree Percentage by State (Ordered by SAT Score)”) merges the data sets “sat_data” and “degree_percent” to show the population 25 and over in each state with a bachelor’s degree by percentage, but the states are ordered in lowest to highest average SAT score. This is attempting to answer my question surrounding the relationship and causality between the percentage of the population with a bachelor’s degree and the average SAT score statewide. The relationship of whether a state with a higher percentage of the population having a bachelor’s degree or greater produces a higher average statewide SAT score and vice versa? The bachelor’s degree percentage is shown along the y-axis, not the SAT scores. The SAT scores are ranked from lowest to highest (left to right) along the x-axis, with the corresponding degree percentage for the state on the y-axis of the bar plot. This presents a clearer plot representation of how the SAT scores and degree percentages relate. Second, the regression outputs is shown, with the intercept value, the main coefficient of interest value, p-value, standard error, and more using the “model summary” code to neatly format a table for the regression (“fit”).
  The bar plot shows that there is not a mutual steady increase. A line of regression was added to estimate the line of best fit and see if a slope was visible. The line shows a slight decrease in the degree percentage as the average SAT score increases, refuting my hypothesis. The main coefficient of interest is shown numerically by my regression table, which shares a -0.022. This means as the State SAT score increases, the percentage of the population with a bachelor’s degree actually lowers, and vice versa, which is the opposite of what I expected. There is statistical significance for the intercept but not for the estimated coefficient in question. This is because the associated p-value is 0.435, a greater value than 0.05, which is a common significance level, meaning it is not statistically significant at that level. Thus, there is less support to be able to reject the null hypothesis with the coefficient equal to zero. Further, from this, I do not believe the coefficient to represent a causal effect. Although the standard error (0.028) is on the smaller side, meaning more precision, the issue of statistical significance from the higher p-value makes a causal effect difficult to support. There are also concerns for bias, confounding or omitted variables, that make basing causality on regression alone a challenge.  
  

Conclusion section:

  As stated above, bar plot created (“Population 25 and Over with a Bachelor’s Degree Percentage by State (Ordered by SAT Score)”) shows the population 25 and over in each state with a bachelor’s degree by percentage, but the states are ordered by lowest to highest average SAT score. The results show a slight decrease of 0.022 in the degree percentage as the average SAT score increases, refuting my hypothesis. From my hypothesis, my results would have expected to show an increase in the degree percentages as the SAT scores increase by state, however, this is not what the results have shown. Some limitations that may have affected my results are that SAT scores from 2023 may not be a reflection of the direction the state is heading in, as the states with lower SAT scores from 2023 may, in the future, have higher degree percentages that could relate to higher SAT scores but that is not discernible yet. Thus, a long-term observational study could be helpful if the time and resources were available. Generalization could also be a specific limitation and threat from not acknowledging the difference in educational systems within states which may lead to higher scores as a result. Thus, the analysis could be improved by looking into state public school educational rankings and the institutions in place that can be a factor in the score an individual might receive. Further, sampling bias could be a threat. This is because the argument that parents with bachelor’s degrees are more expected to have children that have higher SAT scores from them potentially spending more money towards SAT programs or simply having their children take the exam depends on narrowing the sample selection to individuals over 25 with a bachelor’s degree and at least one child, but this study did not exclusively investigate parents. Also, the SAT is mandatory in some states to graduate from high school, but not all. This means that certain states may have more positively skewed scores simply because more individuals took the SAT. Thus, if more data and time was readily available, these potential sampling biases could be accounted for in an easier manner. 
